sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-02 doc/paper.Rmd
Edit; add Fig 1 caption

<!-- # For submission to MEE as an Application paper -->

@@

```{=html}
<!--

Application: Short (no more than 3000-4000 words total) descriptions of new software that are intended to describe and promote the software as well as act as a citeable resource for developers.
Application papers are made free to access for all readers upon publication in order to encourage uptake of the methodologies they describe.
Uploading a package to a site such as [**CRAN**](http://cran.r-project.org/) or [**sourceforge** ](http://sourceforge.net/)in advance is not considered prior publication and will not hinder your submission to MEE.
View some example Applications [**here**](https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/features/applicationpapers).

<https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/features/applicationpapers>

-   Single column, double line spaced

-   Within the word count (no more than 7000-8000 words for Standard Articles, 3000-4000 words for Applications and Practical Tools)

-   Continuous line and page numbering throughout

-   Clearly defined manuscript structure as standard: Author details, Abstract (must be numbered according to [**Manuscript Specifications**](https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210x/author-guidelines#MS_spec)), Keywords, Introduction, Materials and Methods, Results, Discussion, Figures and Tables with captions

-   Figures and Tables can be embedded within the text where referenced to facilitate reviewing

-   Author Contributions ([[if \>1 author -- see Authorship Policy]{.ul}](https://besjournals.onlinelibrary.wiley.com/hub/editorial-policies#authorship)).

-   Statement of where you *intend* to [**archive your data**](https://besjournals.onlinelibrary.wiley.com/hub/data_archiving_policy)

Title Page should include:

A concise and informative title. Do not include the authorities for taxonomic names.
A list of all authors' names with names and addresses of Institutions.
The name, address and e-mail address of the correspondence author.
A running headline of not more than 45 characters.

Abstract

The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:

Key-words: A list in alphabetical order not exceeding eight words or short phrases. The most important key-words should appear in the title and the abstract as well as the key-word list. More advice on selecting good keywords can be found here.

Introduction

This should state the reason for doing the work, the nature of the hypothesis or hypotheses under consideration, and should outline the essential background.

Materials and Methods

Include sufficient details for the work to be repeated. Where specific equipment and materials are named, the manufacturer’s details (name, city and country) should be given so that readers can trace specifications by contacting the manufacturer. Where commercially available software has been used, details of the supplier should be given in brackets or the reference given in full in the reference list. Do not describe or refer to commonplace statistical tests in this section but allude to them briefly in Results.

Results

State the results and draw attention in the text to important details shown in tables and figures.

Discussion

This should point out the significance of the results in relation to the reasons for doing the work, and place them in the context of other work.

Conclusions (optional)

Acknowledgements (optional)

A brief statement acknowledging collaborators and research assistants who do not meet the criteria for authorship described above, or acknowledging funding sources, providing relevant permit numbers (including institutional animal use permits), or giving recognition to nature reserves or other organizations that made the work possible.

Conflict of Interest statement (required for all article types)

-->
```

@@

@@

By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in sdmTMB allow them to be modelled as a random walk or first-order autoregressive, AR(1), process.
We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000B].

>>>>

@@

@@

(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
An sdmTMB model can be built from any combination of these components.
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and 'group', respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components.
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

```{r diagram, fig.cap="(ref:diagram-fig)"}
if (file.exists(here::here("figs/diagram.pdf"))) {
  knitr::include_graphics(here::here("figs/diagram.pdf"))
} else {
  plot(1, type = "n", main = "figs/diagram.pdf missing; will be synced when complete")
}
```

@@

By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in sdmTMB allow them to be modelled as a random walk or first-order autoregressive, AR(1), process.
We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000].
